---
title: "Anthropometric outcomes — BMI Analysis"
subtitle: "Family 1: BMI "
author: "Your name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
    fig-cap-location: bottom
execute:
  echo: true
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    fig.align: center
---

```{r}
cat("Working directory:", getwd(), "\n")
cat("_targets dir exists:", dir.exists("_targets"), "\n")
cat("_targets.R exists:", file.exists("_targets.R"), "\n")
cat("Files here:", paste(head(list.files()), collapse = ", "), "\n")

# Libraries 
library(targets)
library(patchwork)
library(tidyverse)
library(stanviz)

# Functions 
source("R/causal_estimates.R")

# Load the cleaned data from the pipeline and models 
tar_load(data_clean)

## Normal Model
tar_load(stage_1_model_A_normal_pp)
tar_load(stage_1_model_A_normal) 

## Skew Normal Model
tar_load(stage_1_model_A_skewnormal_pp)
tar_load(stage_1_model_A_skewnormal) 

## Student T Model
tar_load(stage_1_model_A_tnormal_pp)
tar_load(stage_1_model_A_tnormal) 


### Additinal Functions #### 
composite <- function(items, sign) {
  
  # Z-score each column
  z_items <- scale(items)
  
  # Weighted composite
  acc <- z_items %*% sign
  
  # Final composite divided by number of items
  comp <- as.vector(acc / ncol(items))
  
  # Z-score the final result
  comp_z <- as.vector(scale(comp))
  
  return(comp_z)
}

```


# MCMC Sampler Diagnostics 

## Trace Plots

### Normal Model 
```{r}
normal_model <- plot_trace(
  model = stage_1_model_A_normal,
  pars  = c("beta_treatment","alpha","sigma")
)
normal_model

```

### Skew Normal Model 
```{r}
skew_model <- plot_trace(
  model = stage_1_model_A_skewnormal,
  pars  = c("beta_treatment","alpha_skew","omega")
)
skew_model
```

### T Student Model 
```{r}
t_model <- plot_trace(
  model = stage_1_model_A_tnormal,
  pars  = c("beta_treatment","alpha","sigma")
)
t_model
```

## Rhats and ESS

### Normal Model 
```{r}
normal_r_hat <- plot_rhat(model = stage_1_model_A_normal)
normal_ess_bulk <- plot_ess(model = stage_1_model_A_normal, type = "bulk")
normal_ess_tail <- plot_ess(model = stage_1_model_A_normal, type = "tail")

normal_r_hat / (normal_ess_bulk + normal_ess_tail) + 
  plot_annotation(
    title = "Rhat and Effective Sample Size Diagnostics",
    theme = theme(
      plot.title = element_text(size = 20)
    ))
```

### Skew Normal 
```{r}
skew_normal_r_hat <- plot_rhat(model = stage_1_model_A_skewnormal)
skew_normal_ess_bulk <- plot_ess(model = stage_1_model_A_skewnormal, type = "bulk")
skew_normal_ess_tail <- plot_ess(model = stage_1_model_A_skewnormal, type = "tail")

skew_normal_r_hat / (skew_normal_ess_bulk + skew_normal_ess_tail) + 
  plot_annotation(
    title = "Rhat and Effective Sample Size Diagnostics",
    theme = theme(
      plot.title = element_text(size = 20)
    ))
```

### Student T 
```{r}
t_normal_r_hat <- plot_rhat(model = stage_1_model_A_tnormal)
t_normal_ess_bulk <- plot_ess(model = stage_1_model_A_tnormal, type = "bulk")
t_normal_ess_tail <- plot_ess(model = stage_1_model_A_tnormal, type = "tail")

t_normal_r_hat / (t_normal_ess_bulk + t_normal_ess_tail) + 
  plot_annotation(
    title = "Rhat and Effective Sample Size Diagnostics",
    theme = theme(
      plot.title = element_text(size = 20)
    ))
```

# Prior Predictive Checks 

````{r}
## Construct the Mass composite
obs_y <- composite(items = as.matrix(data_clean[,c("bmi_post", "weight_post")]), sign = c(1, 1))
````

## Prior Density Plot 

````{r}
normal_prior_dens <- plot_prior_predictive_dens(
  model = stage_1_model_A_normal_pp,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  n_draws       = 100,
  outcome_label = "Mass Post" 
)
skew_normal_prior_dens <- plot_prior_predictive_dens(
  model = stage_1_model_A_skewnormal_pp,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  n_draws       = 100,
  outcome_label = "Mass Post" 
)
t_normal_prior_dens <- plot_prior_predictive_dens(
  model = stage_1_model_A_tnormal,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  n_draws       = 100,
  outcome_label = "Mass Post" 
)
(normal_prior_dens / normal_prior_dens / t_normal_prior_dens) +
  plot_annotation(
    title = "Compare Prior Predictive Distributions across models",
    theme = theme(
      plot.title = element_text(size = 20)
    )
  )
````


## Prior Predictive Stats

### Normal Model 
````{r}
normal_mean <- plot_prior_predictive_stat(
  model    = stage_1_model_A_normal_pp,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  stat     = "mean",
  outcome_label =  "Mass Post"
)

normal_sd <- plot_prior_predictive_stat(
  model    = stage_1_model_A_normal_pp,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  stat     = "sd",
  outcome_label =  "Mass Post"
)

normal_max <- plot_prior_predictive_stat(
  model    = stage_1_model_A_normal_pp,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  stat     = "max",
  outcome_label =  "Mass Post"
)

(normal_mean + normal_sd) / normal_max +
  plot_annotation(
    title = "Normal Model Prior Predictive Statistics",
    subtitle = "Use for Calibration of the Priors",
    theme = theme(
      plot.title = element_text(size = 20, face = "bold"),
      plot.subtitle = element_text(size = 14)
    )
  )
````

### Skew Normal Model 
````{r}
skew_normal_mean <- plot_prior_predictive_stat(
  model    = stage_1_model_A_skewnormal_pp,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  stat     = "mean",
  outcome_label =  "Mass Post"
)

skew_normal_sd <- plot_prior_predictive_stat(
  model    = stage_1_model_A_skewnormal_pp,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  stat     = "sd",
  outcome_label =  "Mass Post"
)

skew_normal_max <- plot_prior_predictive_stat(
  model    = stage_1_model_A_skewnormal_pp,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  stat     = "max",
  outcome_label =  "Mass Post"
)

(skew_normal_mean + skew_normal_sd) / skew_normal_max +
  plot_annotation(
    title = "Skew Normal Model Prior Predictive Statistics",
    subtitle = "Use for Calibration of the Priors",
    theme = theme(
      plot.title = element_text(size = 20, face = "bold"),
      plot.subtitle = element_text(size = 14)
    )
  )
````

### Student T Model 
````{r}
t_normal_mean <- plot_prior_predictive_stat(
  model    = stage_1_model_A_tnormal_pp,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  stat     = "mean",
  outcome_label =  "Mass Post"
)

t_normal_sd <- plot_prior_predictive_stat(
  model    = stage_1_model_A_tnormal_pp,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  stat     = "sd",
  outcome_label =  "Mass Post"
)

t_normal_max <- plot_prior_predictive_stat(
  model    = stage_1_model_A_tnormal_pp,
  yrep_var = "mass_post_rep",
  y        = obs_y,
  stat     = "max",
  outcome_label =  "Mass Post"
)

(t_normal_mean + t_normal_sd) / t_normal_max +
  plot_annotation(
    title = "Student T Model Prior Predictive Statistics",
    subtitle = "Use for Calibration of the Priors",
    theme = theme(
      plot.title = element_text(size = 20, face = "bold"),
      plot.subtitle = element_text(size = 14)
    )
  )
````

# Posterior predictive Checks 

## Posterior Predictive Density 

### Normal Model 
```{r}
normal_post <- plot_ppc_dens(
  model = stage_1_model_A_normal,
  y        =  obs_y,
  yrep_var = "mass_post_rep",
  outcome_label = "Mass Post"
)

(normal_post + normal_prior_dens) + 
  plot_annotation(
    title = "Comparison between Prior and Posterior for Normal Model",
    theme = theme(
      plot.title = element_text(size = 20)
    )
  ) 

``` 

### Skew Normal Model 
```{r}
skew_normal_post <- plot_ppc_dens(
  model = stage_1_model_A_skewnormal,
  y        =  obs_y,
  yrep_var = "mass_post_rep",
  outcome_label = "Mass Post"
)

(skew_normal_post + skew_normal_prior_dens) + 
  plot_annotation(
    title = "Comparison between Prior and Posterior for Skew Normal Model",
    theme = theme(
      plot.title = element_text(size = 20)
    )
  ) 

``` 

### Skew Normal Model 
```{r}
t_normal_post <- plot_ppc_dens(
  model = stage_1_model_A_tnormal,
  y        =  obs_y,
  yrep_var = "mass_post_rep",
  outcome_label = "Mass Post"
)

(t_normal_post + t_normal_prior_dens) + 
  plot_annotation(
    title = "Comparison between Prior and Posterior for Student T Model",
    theme = theme(
      plot.title = element_text(size = 20)
    )
  ) 

``` 


## Posterior Predictive Statistics

### Normal 
```{r}
normal_post_mean <- plot_ppc_stat(
  model = stage_1_model_A_normal,
  y        = obs_y,
  yrep_var = "mass_post_rep",
  stat     = "mean"    
)

normal_post_sd <- plot_ppc_stat(
  model = stage_1_model_A_normal,
  y        = obs_y,
  yrep_var = "mass_post_rep",
  stat     = "sd"    
)
(normal_post_mean + normal_post_sd) +
  plot_annotation(
    title = "Posterior Predictive Statiscts for Normal Model",
    theme = theme(
      plot.title = element_text(size = 20)
    )
  )
```

### Skew Normal 
```{r}
skew_normal_post_mean <- plot_ppc_stat(
  model = stage_1_model_A_skewnormal,
  y        = obs_y,
  yrep_var = "mass_post_rep",
  stat     = "mean"    
)

skew_normal_normal_post_sd <- plot_ppc_stat(
  model = stage_1_model_A_skewnormal,
  y        = obs_y,
  yrep_var = "mass_post_rep",
  stat     = "sd"    
)
(skew_normal_post_mean + skew_normal_normal_post_sd) +
  plot_annotation(
    title = "Posterior Predictive Statiscts for Skew Normal Model",
    theme = theme(
      plot.title = element_text(size = 20)
    )
  )
```

### Student T 
```{r}
t_normal_post_mean <- plot_ppc_stat(
  model = stage_1_model_A_tnormal,
  y        = obs_y,
  yrep_var = "mass_post_rep",
  stat     = "mean"    
)

t_normal_normal_post_sd <- plot_ppc_stat(
  model = stage_1_model_A_tnormal,
  y        = obs_y,
  yrep_var = "mass_post_rep",
  stat     = "sd"    
)
(t_normal_post_mean + t_normal_normal_post_sd) +
  plot_annotation(
    title = "Posterior Predictive Statiscts for Student T Model",
    theme = theme(
      plot.title = element_text(size = 20)
    )
  )
```


## Per-observation predictive intervals
```{r}
normal_ppc_int <- plot_ppc_intervals(
  model = stage_1_model_A_normal,
  y        = obs_y,
  yrep_var = "mass_post_rep",
  outcome_label = "Mass Post"
)
skew_normal_ppc_int <- plot_ppc_intervals(
  model = stage_1_model_A_skewnormal,
  y        = obs_y,
  yrep_var = "mass_post_rep",
  outcome_label = "Mass Post"
)
t_normal_ppc_int <- plot_ppc_intervals(
  model = stage_1_model_A_tnormal,
  y        = obs_y,
  yrep_var = "mass_post_rep",
  outcome_label = "Mass Post"
)
(normal_ppc_int / skew_normal_ppc_int / t_normal_ppc_int) +
  plot_annotation(
    title = "Per-observation predictive intervals Comparison",
    theme = theme(
      plot.title = element_text(size = 20)
    ) 
  )
```

## Predictive error structure across fitted values and PIT calibration histogram

### Normal 
```{r}
normal_pred_error <- plot_ppc_error_scatter(
  model = stage_1_model_A_normal,
  y     = obs_y,
  yrep_var = "mass_post_rep"
)

normal_pit <- plot_ppc_pit(model = stage_1_model_A_normal)

(normal_pred_error + normal_pit) + 
  plot_annotation(
    title = "Predictive error and PIT calibration Normal Model",
    theme = theme(
      plot.title = element_text(size = 20)
    )
  )

```

### Skew Normal 
```{r}
skew_normal_pred_error <- plot_ppc_error_scatter(
  model = stage_1_model_A_skewnormal,
  y     = obs_y,
  yrep_var = "mass_post_rep"
)

skew_normal_pit <- plot_ppc_pit(model = stage_1_model_A_skewnormal)

(skew_normal_pred_error + skew_normal_pit) + 
  plot_annotation(
    title = "Predictive error and PIT calibration Skew Normal Model",
    theme = theme(
      plot.title = element_text(size = 20)
    )
  )
```

### Student T  
```{r}
t_normal_pred_error <- plot_ppc_error_scatter(
  model = stage_1_model_A_tnormal,
  y     = obs_y,
  yrep_var = "mass_post_rep"
)

t_normal_pit <- plot_ppc_pit(model = stage_1_model_A_tnormal)

(t_normal_pred_error + t_normal_pit) + 
  plot_annotation(
    title = "Predictive error and PIT calibration Student T Model",
    theme = theme(
      plot.title = element_text(size = 20)
    )
  )
```

# Parameter posteriors

## Skew Normal Forest Coeff Plot and Normal Comparison
```{r}
skew_forest <- plot_coef_forest(
  model = stage_1_model_A_skewnormal,
  pars = c("alpha_skew", "alpha","beta_treatment","omega","beta_mass_pre")
)
skew_forest

normal_forest <- plot_coef_forest(
  model = stage_1_model_A_normal,
  pars = c("alpha","beta_treatment","sigma","beta_mass_pre")
)
(skew_forest / normal_forest) +
  plot_annotation(
    title = "Model Coeff Comaprisons Between Skew Normal and Normal Model",
    theme = theme(
      plot.title = element_text(size = 20)
    )
  ) 
```

## Skew Normal Halfeye densities
```{r}
skew_dens_coef <- plot_param_halfeye(
  model = stage_1_model_A_skewnormal,
  pars = c("alpha_skew", "alpha","beta_treatment","omega","beta_mass_pre")
)
skew_dens_coef
```

## Normal Halfeye densities
```{r}
normal_dens_coef <- plot_param_halfeye(
  model = stage_1_model_A_normal,
  pars = c("alpha","beta_treatment","sigma","beta_mass_pre")
)
normal_dens_coef
```

# Model comparison

## Skew Normal Vs Normal Model 
```{r}
comp_loo_normal <- compute_loo(model = stage_1_model_A_normal)
comp_loo_skew <- compute_loo(model = stage_1_model_A_skewnormal)

plot_loo_compare(list(normal = comp_loo_normal, skew = comp_loo_skew))
```



