---
title: "Anthropometric outcomes — BMI Analysis"
subtitle: "Family 1: BMI "
author: "Your name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
    fig-cap-location: bottom
execute:
  echo: true
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    fig.align: center
---

```{r}
cat("Working directory:", getwd(), "\n")
cat("_targets dir exists:", dir.exists("_targets"), "\n")
cat("_targets.R exists:", file.exists("_targets.R"), "\n")
cat("Files here:", paste(head(list.files()), collapse = ", "), "\n")

# Libraries 
library(targets)
library(patchwork)
library(tidyverse)
library(stanviz)

# Functions 
source("R/causal_estimates.R")

# Load the cleaned data from the pipeline and models 
tar_load(data_clean)

## Model B1: Diet ~ Treatment + Diet_pre
tar_load(stage_2_model_B1)
tar_load(stage_2_model_B1_pp) 

## Model B2: Activity  ~ Treatment + Activity_pre
tar_load(stage_2_model_B2)
tar_load(stage_2_model_B2_pp) 

## Model B3: SelfCare  ~ Treatment + SelfCare_pre
tar_load(stage_2_model_B3)
tar_load(stage_2_model_B3_pp) 


### Additinal Functions #### 
composite <- function(items, sign) {
  
  # Z-score each column
  z_items <- scale(items)
  
  # Weighted composite
  acc <- z_items %*% sign
  
  # Final composite divided by number of items
  comp <- as.vector(acc / ncol(items))
  
  # Z-score the final result
  comp_z <- as.vector(scale(comp))
  
  return(comp_z)
}

```

# Model B1: Diet ~ Treatment + Diet_pre

```{r}
y_obs <- as.vector(scale(data_clean$diet_score_post))
```

## MCMC Sampler Diagnostocs

### Trace Rank Plots
```{r}
plot_rank(
    model = stage_2_model_B1,
    pars  = c("beta_treatment", "beta_diet_pre", "alpha", "sigma")
)
``` 

### Rhats and ESS
```{r}
b1_rhat     <- plot_rhat(model = stage_2_model_B1)
b1_ess_bulk <- plot_ess(model  = stage_2_model_B1, type = "bulk")
b1_ess_tail <- plot_ess(model  = stage_2_model_B1, type = "tail")

(b1_ess_bulk + b1_ess_tail) / b1_rhat + 
    plot_annotation(
        title = "Rhat and Effective Sample Size Tail and Bulk",
        theme = theme(
            plot.title = element_text(size = 20)
        )
    )

```

## Prior Predictive Checks

### Prior Predictive Distribution 
```{r}
plot_prior_predictive_dens(
    model = stage_2_model_B1_pp,
    yrep_var = "diet_post_rep",
    y        = y_obs,
    n_draws = 100,
    outcome_label = "Diet Score Post-Intervention"
)
```

### Prior Predictive Stats
```{r}
mean_stat <- plot_prior_predictive_stat(
    model = stage_2_model_B1_pp,
    yrep_var = "diet_post_rep",
    y        = y_obs,
    stat     = "mean", 
    outcome_label = "Prior Predictive Mean"
)
sd_stat <- plot_prior_predictive_stat(
    model = stage_2_model_B1_pp,
    yrep_var = "diet_post_rep",
    y        = y_obs,
    stat     = "sd", 
    outcome_label = "Prior Predictive SD"
)

(mean_stat + sd_stat) + 
    plot_annotation(
        title = "Prior Predictive Mean and SD",
        theme = theme(
            plot.title = element_text(size = 20)
        )
    )
```

## Posterior Predictive Checks

### Posterior Predictive Density 
```{r}
plot_ppc_dens(
    model = stage_2_model_B1,
    y = y_obs,
    yrep_var = "diet_post_rep",
    outcome_label = "Post-Intervention Diet Score"
)
```

### Posterior Predictive Stats
```{r}
post_mean <- plot_ppc_stat(
    model = stage_2_model_B1,
    y = y_obs,
    yrep_var = "diet_post_rep",
    stat     = "mean" 
)
post_sd <- plot_ppc_stat(
    model = stage_2_model_B1,
    y = y_obs,
    yrep_var = "diet_post_rep",
    stat     = "sd"
)
post_intervals <- plot_ppc_intervals(
    model = stage_2_model_B1,
    y = y_obs,
    yrep_var = "diet_post_rep",
    outcome_label = "Post Intervention Diet Score"
)

(post_mean + post_sd) / post_intervals+ 
    plot_annotation(
        title = "Posterior Predictive Mean and SD",
        theme = theme(
            plot.title = element_text(size = 20)
        )
    )

```

### Posterior Errors
```{r}
post_err <- plot_ppc_error_scatter(
    model = stage_2_model_B1,
    y = y_obs,
    yrep_var = "diet_post_rep"
)
pit_calid <- plot_ppc_pit(model = stage_2_model_B1, pit_var = "pit" )

(post_err + pit_calid) + 
    plot_annotation(
        title = "Posterior Predictive Error Checks",
        theme = theme(
            plot.title = element_text(size = 20)
        )
    )
```

# Model B2: Activity  ~ Treatment + Activity_pre

```{r}
y_obs <- composite(items = )
```

## MCMC Sampler Diagnostocs

### Trace Rank Plots
```{r}
plot_rank(
    model = stage_2_model_B1,
    pars  = c("beta_treatment", "beta_diet_pre", "alpha", "sigma")
)
``` 

### Rhats and ESS
```{r}
b1_rhat     <- plot_rhat(model = stage_2_model_B1)
b1_ess_bulk <- plot_ess(model  = stage_2_model_B1, type = "bulk")
b1_ess_tail <- plot_ess(model  = stage_2_model_B1, type = "tail")

(b1_ess_bulk + b1_ess_tail) / b1_rhat + 
    plot_annotation(
        title = "Rhat and Effective Sample Size Tail and Bulk",
        theme = theme(
            plot.title = element_text(size = 20)
        )
    )

```

## Prior Predictive Checks

### Prior Predictive Distribution 
```{r}
plot_prior_predictive_dens(
    model = stage_2_model_B1_pp,
    yrep_var = "diet_post_rep",
    y        = y_obs,
    n_draws = 100,
    outcome_label = "Diet Score Post-Intervention"
)
```

### Prior Predictive Stats
```{r}
mean_stat <- plot_prior_predictive_stat(
    model = stage_2_model_B1_pp,
    yrep_var = "diet_post_rep",
    y        = y_obs,
    stat     = "mean", 
    outcome_label = "Prior Predictive Mean"
)
sd_stat <- plot_prior_predictive_stat(
    model = stage_2_model_B1_pp,
    yrep_var = "diet_post_rep",
    y        = y_obs,
    stat     = "sd", 
    outcome_label = "Prior Predictive SD"
)

(mean_stat + sd_stat) + 
    plot_annotation(
        title = "Prior Predictive Mean and SD",
        theme = theme(
            plot.title = element_text(size = 20)
        )
    )
```

## Posterior Predictive Checks

### Posterior Predictive Density 
```{r}
plot_ppc_dens(
    model = stage_2_model_B1,
    y = y_obs,
    yrep_var = "diet_post_rep",
    outcome_label = "Post-Intervention Diet Score"
)
```

### Posterior Predictive Stats
```{r}
post_mean <- plot_ppc_stat(
    model = stage_2_model_B1,
    y = y_obs,
    yrep_var = "diet_post_rep",
    stat     = "mean" 
)
post_sd <- plot_ppc_stat(
    model = stage_2_model_B1,
    y = y_obs,
    yrep_var = "diet_post_rep",
    stat     = "sd"
)
post_intervals <- plot_ppc_intervals(
    model = stage_2_model_B1,
    y = y_obs,
    yrep_var = "diet_post_rep",
    outcome_label = "Post Intervention Diet Score"
)

(post_mean + post_sd) / post_intervals+ 
    plot_annotation(
        title = "Posterior Predictive Mean and SD",
        theme = theme(
            plot.title = element_text(size = 20)
        )
    )

```

### Posterior Errors
```{r}
post_err <- plot_ppc_error_scatter(
    model = stage_2_model_B1,
    y = y_obs,
    yrep_var = "diet_post_rep"
)
pit_calid <- plot_ppc_pit(model = stage_2_model_B1, pit_var = "pit" )

(post_err + pit_calid) + 
    plot_annotation(
        title = "Posterior Predictive Error Checks",
        theme = theme(
            plot.title = element_text(size = 20)
        )
    )
```